---
title: "Analysing differential localisation experiments with BANDLE"
author:
- name: Oliver M. Crook
package: bandle
abstract: >
  This vignette describes how to analyse a mass-spectrometry based 
  differential localisation experiments using BANDLE. Data should be
  stored as lists of `MSnSet`s. There is also features for quality control and
  visualisation of results. Other vignettes are for convergence and other
  methodology
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Analysing differential localisation experiments with BANDLE}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, MS, MSMS, Proteomics, Metabolomics, Infrastructure, Quantitative}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, message = FALSE, warning = FALSE, echo = FALSE}
options(width = 150)
library("bandle")
library("ggplot2")
library("pRoloc")
library("pRolocdata")
library("MSnbase")
library("pheatmap")
library("viridis")

setStockcol(paste0(getStockcol(),"90")) # see throught colours
```

# A well-defined differential localisation experiment

We are going to generate a differential localisation experiment. They key 
elements are replicates, and a perturbation of interest. There is code within
the `bandle` package to simulate an example experiment. First we will use 
the `pRolocdata` package to obtain a spatial proteomics dataset. This will be
the basis of our simulation which will use boostrapping to generate new datasets.

```{r,}
library("pRolocdata")
data("tan2009r1")
```

```{r exampledata,}
plot2D(tan2009r1,
       main = "example spatial proteomics datasets", grid = FALSE)
addLegend(tan2009r1, where = "topleft", cex = 0.7)
```
The following code chuck simulates a differential localisation experiment.
It will generate `numRep/2` of each a control and treatment condition. We
will also simulate relocalisations for `numDyn` proteins.
```{r simulatedataset,}
set.seed(1)
tansim <- sim_dynamic(object = tan2009r1,
                      numRep = 6,
                      numDyn = 100)

```

```{r visualisationdata, out = FALSE, message = FALSE, echo=FALSE, fig.width=15, fig.height=15}
par(mfrow = c(3,2))
out <- lapply(tansim$lopitrep, function(z) plot2D(z, grid = FALSE))
```
For understanding, exploring and visualizing individual spatial proteomics
experiments, see the vignettes in `pRoloc` and `MSnbase` packages.  

```{r,}
tansim$lopitrep[[1]]
```
# Preparing for BANDLE analysis

The main function of the bandle package is `bandle`, this uses a complex
model to analyse the data. Markov-Chain Monte-Carlo (MCMC) is used to 
sample the posterior distribution of parameters and latent variables.
From which statistics of interest can be computed. Here we only run a few iterations
for brevity but typically one needs to run thousands of iterations to ensure
convergence, as well as multiple parallel chains.

First, we need to fit non-parametric regression functions to the markers profiles,
upon which we place our analysis. This uses Gaussian processes. The fitting uses
some default penalised complexity priors, which work well. However, these
can be altered, which is demonstrated in the next code chunk


```{r,}
par(mfrow = c(3,4))
gpParams <- lapply(tansim$lopitrep, function(x) fitGPmaternPC(x))
```
The prior needs to form a `K*3` matrix, which gives the prior the length-scale
amplitude and standard deviation parameters. Increasing these values, increases
the shrinkage. For more details see the bandle paper.

```{r,}
K <- length(getMarkerClasses(tansim$lopitrep[[1]]))
pc_prior <- matrix(NA, ncol = 3, K)
pc_prior[seq.int(1:K), ] <- matrix(rep(c(10, 60, 250),
                                       each = K), ncol = 3)
```


```{r,}
par(mfrow = c(3,4))
gpParams <- lapply(tansim$lopitrep,
                   function(x) fitGPmaternPC(x, hyppar = pc_prior))
```

# Setting the prior on the weights

The next step is to set up the matrix Dirichlet prior on the mixing weights. These
weights are defined across datasets so these are slightly different to mixture
weights in usual mixture models. The $(i,j)^{th}$ entry is the prior probability
that a protein localises to organelle $i$ in the control and $j$ in the treatment.
This mean that off-diagonal terms have a different interpretation to diagonal terms.
Since we expect relocalisation to be rare, off-diagonal terms should be small.
The following functions help set up the priors and how to interpret them. The
parameter `q` allow us to check the prior probability that more than `q`
differential localisations are expected.

```{r,}
dirPrior = diag(rep(1, K)) + matrix(0.001, nrow = K, ncol = K)
predDirPrior <- prior_pred_dir(object = tansim$lopitrep[[1]],
                               dirPrior = dirPrior,
                               q = 15)
```

The mean number of relocalisation is small:
```{r,}
predDirPrior$meannotAlloc
```

The prior probability that more than `q` differential localisations are
expected is small
```{r,}
predDirPrior$tailnotAlloc
```

The full prior predictive can be visualised as histogram. The prior probability
that proteins are allocated to different components between datasets concentrates
around 0.

```{r,}
hist(predDirPrior$priornotAlloc, col = getStockcol()[1])
```

# The bandle function

We are now ready to run the main bandle function. Remember to carefully
select the datasets and replicates that define the control and treatment. 
Furthermore, iterations are typically $1000$s, a few are run here for the 
convenience of the vignette.
```{r,}
control <- tansim$lopitrep[1:3]
treatment <- tansim$lopitrep[4:6]
bandleres <- bandle(objectCond1 = control,
                    objectCond2 = treatment,
                    numIter = 50, # usually 10,000
                    burnin = 5, # usually 5,000
                    thin = 1, # usually 20
                    gpParams = gpParams,
                    pcPrior = pc_prior,
                    numChains = 1, # usually >=4
                    dirPrior = dirPrior)

```
The bandle function generates an object of class `bandleParams`. The `show`
method indicates the number of parallel chains that were run, this should
typically be greater than 4.
```{r,}
bandleres
```

# The summary slot

Currently, the summary slots of the `bandleres` object are empty. The `summaries`
function accesses them.
```{r,}
summaries(bandleres)
```

These can be populated as follows
```{r,}
bandleres <- bandleProcess(bandleres)
```

These slot have now been populated
```{r,}
summary(summaries(bandleres))
```

The `posteriorEstimates` slot gives posterior quantities of interest for different
proteins. The object is of length 2, 1 for control and 1 for treatment.
```{r,}
length(summaries(bandleres))
```

One quantity of interest is the protein allocations, which we can plot in a barplot
```{r,}
par(mfrow = c(1, 2))

pe1 <- summaries(bandleres)[[1]]@posteriorEstimates
barplot(table(pe1$bandle.allocation), col = getStockcol()[2], las = 2, main = "Protein allocation: control")

pe2 <- summaries(bandleres)[[2]]@posteriorEstimates
barplot(table(pe2$bandle.allocation), col = getStockcol()[2], las = 2, main = "Protein allocation: treatment")
```

# Differential localisation probability

The differential localization probability is also stored here, which gives
the proteins that are most likely to differentially localised. The rank plot
is a good visualisation. Indicating that most proteins are not differentially 
localised and there are a few confident differentially localised protiens.

```{r,}
diffloc_probs <- pe1$bandle.differential.localisation
plot(diffloc_probs[order(diffloc_probs, decreasing = TRUE)],
     col = getStockcol()[3], pch = 19, ylab = "Probability", 
     xlab = "Rank", main = "Differential localisation rank plot")
```
We can examine the top `n` proteins (here 100) and produce bootstrap estimates 
of the uncertainty (note here the uncertainty is likely to be underestimated
as we did not produce many mcmc samples). These can be visualised as ranked
boxplots

```{r,}
set.seed(1)
bt <- bootstrapdiffLocprob(params = bandleres, top = 100)
boxplot(t(bt), col = getStockcol()[5],
        las = 2, ylab = "Probability", ylim = c(0, 1),
        main = "Differential localisation probability plot (top 100 proteins)")
```
Let see how many proteins in the top 100 match those we had simulated relocalisations
for. 
```{r,}
sum(tansim$perm1_names %in% rownames(bt))
sum(!(tansim$perm1_names %in% rownames(bt)))
```

# Visualising relocalisation

We can visualise the changes in localisation between conditions on an alluvial 
(river) plot using the `riverplot` function
```{r, warning=FALSE, message=FALSE}
riverplot(bandleres)
```


# Allocation probabilities 

The allocation probabilities are stored in the `tagm.joint` slot. These could
be visualised in a heatmap
```{r,}
bjoint_control <- summaries(bandleres)[[1]]@bandle.joint
pheatmap(bjoint_control, cluster_cols = FALSE, color = viridis(n = 25))
```
```{r,}
bjoint_treatment <- summaries(bandleres)[[2]]@bandle.joint
pheatmap(bjoint_treatment, cluster_cols = FALSE, color = viridis(n = 25))
```

# Session information

All software and respective versions used to produce this document are listed below.
```{r,}
sessionInfo()
```
