---
title: "Analysing differential localisation experiments with BANDLE"
author:
- name: Oliver M. Crook
package: bandle
abstract: >
  This vignette describes how to analysis a mass-spectrometry based 
  differential localisation experiments using BANDLE. Data should be
  stored as lists of `MSnSet`. There is also features for quality control and
  visualisation of results
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Analysing differential localisation experiments with BANDLE}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, MS, MSMS, Proteomics, Metabolomics, Infrastructure, Quantitative}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, message = FALSE, warning = FALSE, echo = FALSE}
library("bandle")
library("ggplot2")
library("pRoloc")
library("pRolocdata")
library("MSnbase")

setStockcol(paste0(getStockcol(),"90")) # see throught colours
```

# A well-defined differential localisation experiment

We are going to generate a differential localisation experiment. They key 
elements are replicates, and a perturbation of interest. There is code within
the `bandle` package to simulate an example experiment. First we will use 
the `pRolocdata` package to obtain a spatial proteomics dataset. This will be
the basis of our simulation which will use boostrapping to generate new datasets.

```{r,}
data("tan2009r1")
```

```{r exampledata,}
plot2D(tan2009r1, main = "example spatial proteomics datasets", grid = FALSE)
addLegend(tan2009r1, where = "topleft", cex = 0.7)
```
The following code chuck simulates a differential localisation experiment.
It will generate `numRep/2` of each a control and treatment condition. We
will also simulate relocalisations for `numDyn` proteins.
```{r simulatedataset,}

tansim <- sim_dynamic(object = tan2009r1,
                      numRep = 6,
                      numDyn = 20)

```

```{r visualisationdata, out = FALSE, message = FALSE, print = FALSE, echo=FALSE}
par(mfrow = c(3,2))
out <- lapply(tansim$lopitrep, function(z) plot2D(z, grid = FALSE))
```
For understanding, exploring and visualizing individual spatial proteomics
experiments, see vignettes in `pRoloc` and `MSnbase` package.  

```{r,}
tansim$lopitrep[[1]]
```
The main function of the bandle package is `bandle`, this uses a complex
model to analyse the data. Markov-Chain Monte-Carlo (MCMC) is used to 
sample the posterior distribution of parameters and latent variables.
From which statistics of interest can be computed. Here we only run a few iteration
for brevity but typically one need to run thousands of iterations to ensure
convergence, as well as multiple parallel chains.

First, we need to fit non-parametric regression functions to the markers profiles,
upon which we place our analysis. This uses Gaussian processes. The fitting uses
some default penalised complexity priors, which work well. However, these
can be altered, which is demonstrated in the next code chunk


```{r,}
par(mfrow = c(3,4))
gpParams <- lapply(tansim$lopitrep, function(x) fitGPmaternPC(x))
```
To prior need to form a `K*3` matrix, which give the prior for the length-scale
amplitude and standard deviation parameters. Increasing these values, increases
the shrinkage. For more details see the bandle paper.

```{r,}
K <- length(getMarkerClasses(tansim$lopitrep[[1]]))
pc_prior <- matrix(NA, ncol = 3, K)
pc_prior[seq.int(1:K), ] <- matrix(rep(c(10, 60, 250), each = K), ncol = 3)
```


```{r,}
par(mfrow = c(3,4))
gpParams <- lapply(tansim$lopitrep, function(x) fitGPmaternPC(x, hyppar = pc_prior))
```
The next step is to set up the matrix dirichlet prior on the mixing weights. These
weights are defined across datasets so these are slightly different to mixture
weights in usual mixture models. The $(i,j)$ entry is the prior probability
that a protein localises to organelle $i$ in the control and $j$ in the treatment.
This mean that off-diagonal terms have a different interpretation to diagonal terms.
Since we expect relocalisation to be rare off-diagonal terms should be small.
The following functions help set up the priors and how to interpret them. The
parameter `q` allow us to check the prior probability that more than `q`
differential localisation are expected.

```{r,}
dirPrior = diag(rep(1, K)) + matrix(0.001, nrow = K, ncol = K)
predDirPrior <- prior_pred_dir(object = tansim$lopitrep[[1]],
                               dirPrior = dirPrior,
                               q = 15)
```

The mean number of relocalisation is small:
```{r,}
predDirPrior$meannotAlloc
```

The prior probability than more than `q` differential localisation are
expected is small
```{r,}
predDirPrior$tailnotAlloc
```

The full prior predictive can be visualised as histogram. The prior probability
that proteins are allocated to different components between datasets concentrates
around 0.

```{r,}
hist(predDirPrior$priornotAlloc, col = getStockcol()[1])
```
We are now ready to run the main bandle function. Remember to carefully
select the dataset and replicates that define the control and treatment. 
Furthermore, iteration are typically $1000$s, a few a run here for the convenience
of the vignette.
```{r,}
control <- tansim$lopitrep[1:3]
treatment <- tansim$lopitrep[4:6]
bandleres <- bandle(objectCond1 = control,
                    objectCond2 = treatment,
                    numIter = 20,
                    burnin = 5, 
                    thin = 1,
                    gpParams = gpParam, 
                    numChains = 1, 
                    dirPrior = dirPrior)

```



