---
title: "Vignette 2: A workflow for analysing differential localisation"
author:
- name: Oliver M. Crook 
  affiliation: Department of Statistics, University of Oxford, UK
- name: Lisa M. Breckels
  affiliation: Cambridge Centre for Proteomics, University of Cambridge, UK
package: bandle
abstract: >
  This vignette describes how to analyse mass-spectrometry based 
  differential localisation experiments using the BANDLE method [@bandle]. 
  Data should be stored as lists of `MSnSet`s. There is also features for 
  quality control and visualisation of results. Please see other vignettes for 
  convergence and other methodology.
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: bandle.bib
vignette: >
  %\VignetteIndexEntry{Analysing differential localisation experiments with BANDLE: Vignette 2}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, MS, MSMS, Proteomics, Metabolomics, Infrastructure, Quantitative}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=300,fig.width=7)
```

```{r env, message = FALSE, warning = FALSE, echo = FALSE}
# options(width = 150)
library("bandle")
library("ggplot2")
library("pRoloc")
library("MSnbase")
library("pheatmap")
library("viridis")
library("dplyr")

setStockcol(paste0(getStockcol(),"90")) # make colours transparent
```

# Introduction

In this vignette we use a real-life biological use-case to demonstrate how to
analyse mass-spectrometry based proteomics data using the Bayesian ANalysis of
Differential Localisation Experiments (BANDLE) method.

# The data

As mentioned in "Vignette 1: Getting Started with BANDLE" data from mass
spectrometry based proteomics methods most commonly yield a matrix of
measurements where we have proteins/peptides/peptide spectrum matches
(PSMs) along the rows, and samples/fractions along the columns. To use `bandle`
the data must be stored as a `MSnSet`, as implemented in the Bioconductor 
`r Biocpkg("MSnbase")` package. Please see the relevant vignettes in 
`r Biocpkg("MSnbase")` for constructing these data containers.

The data used in this vignette has been published in @thplopit and is currently
stored as `MSnSet` instances in the the `r Biocpkg("pRolocdata")` package. We
will load it in the next section.

## Spatialtemporal proteomic profiling of a THP-1 cell line

In this workflow we analyse the data produced by @thplopit. In this experiment
triplicate hyperLOPIT experiments [@Mulvey:2017] were conducted on THP-1 human
leukaemia cells where the samples were analysed and collected (1) when cells
were unstimulated and then (2) following 12 hours stimulation with LPS
(12h-LPS).

In the following code chunk we load 4 of the datasets from the study: 2
replicates of the unstimulated and 2 replicates of the 12h-LPS stimulated
samples. Please note to adhere to Bioconductor vignette build times we only load
2 of the 3 replicates for each condition to demonstrate the BANDLE workflow.

```{r loaddata}
library("pRolocdata")
data("thpLOPIT_unstimulated_rep1_mulvey2021")
data("thpLOPIT_unstimulated_rep3_mulvey2021")
data("thpLOPIT_lps_rep1_mulvey2021")
data("thpLOPIT_lps_rep3_mulvey2021")
```

By typing the names of the datasets we get a `MSnSet` data summary. For
example,

```{r summarydata}
thpLOPIT_unstimulated_rep1_mulvey2021
thpLOPIT_lps_rep1_mulvey2021
```

We see that the datasets `thpLOPIT_unstimulated_rep1_mulvey2021` and
`thpLOPIT_lps_rep1_mulvey2021` contain 5107 and 4879 proteins respectively,
across 20 TMT channels. The data is accessed through different slots of the
`MSnSet` (see `str(thpLOPIT_unstimulated_rep1_mulvey2021)` for all available
slots). The 3 main slots which are used most frequently are those that contain
the quantitation data, the features i.e. PSM/peptide/protein information and the
sample information, and these can be accessed using the functions `exprs`,
`fData`, and `pData`, respectively.

## Preparing the data

To run `bandle` there are a few minimal requirements that the data must fulfill.
Data are required to have: 

- the same number of channels across conditions and replicates
- the same proteins across conditions and replicates
- data must be a `list` of `MSnSet` instances

If we use the `dim` function we see that the datasets we have loaded have the
same number of channels but a different number of proteins per experiment.

```{r datadim}
dim(thpLOPIT_unstimulated_rep1_mulvey2021)
dim(thpLOPIT_unstimulated_rep3_mulvey2021)
dim(thpLOPIT_lps_rep1_mulvey2021)
dim(thpLOPIT_lps_rep3_mulvey2021)
```

We use the function `commonFeatureNames` to extract proteins that are common
across all replicates. This function has a nice side effect which is that it
also wraps the data into a `list`, ready for input into `bandle`.


```{r cmnprots}
thplopit <- commonFeatureNames(c(thpLOPIT_unstimulated_rep1_mulvey2021,  ## unstimulated rep
                                 thpLOPIT_unstimulated_rep3_mulvey2021,  ## unstimulated rep
                                 thpLOPIT_lps_rep1_mulvey2021,           ## 12h-LPS rep
                                 thpLOPIT_lps_rep3_mulvey2021))          ## 12h-LPS rep
```

We now have our list of `MSnSet`s ready for `bandle` with 3727 proteins common
across all 4 replicates/conditions.

```{r listmsnsets}
thplopit
```

We can visualise the data using the `plot2D` function from `pRoloc`

```{r exampledata, fig.height=10, fig.width=10}
## create a character vector of title names for the plots
plot_id <- c("Unstimulated 1st rep", "Unstimulated 2nd rep",
             "12h-LPS 1st rep", "12h-LPS 2nd rep")

## plot the data
par(mfrow = c(2,2))
for (i in seq(thplopit))
    plot2D(thplopit[[i]], main = plot_id[i])
addLegend(thplopit[[4]], where = "topleft", cex = .75)
```

By default the `plot2D` uses principal components analysis (PCA)
for the data transformation. Other options such as t-SNE, kernal
PCA etc. are also available, see `?plot2D` and the `method` argument.
PCA sometimes will randomly flip the axis, because the eigenvectors 
only need to satisfy $||v|| = 1$, which allows a sign flip. You will
notice this is the case for the 3rd plot. If desired you can flip
the axis/change the sign of the PCs by specifying any of the arguments
`mirrorX`, `mirrorY`, `axsSwitch` to TRUE when you call `plot2D`.

# Preparing `bandle`: fitting GPs and setting the priors

As mentioned in the first vignette, `bandle` uses a complex model to analyse the
data. Markov-Chain Monte-Carlo (MCMC) is used to sample the posterior
distribution of parameters and latent variables from which statistics of
interest can be computed. Again, here we only run a few iterations for brevity
but typically one needs to run thousands of iterations to ensure convergence, as
well as multiple parallel chains.

## Fitting Gaussian processes

First, we need to fit non-parametric regression functions to the markers
profiles. We use the `fitGPmaternPC` function using the default penalised
complexity priors (see `?fitGP`), which work well. 


```{r fitgps}
par(mfrow = c(3,4))
gpParams <- lapply(thplopit, function(x) fitGPmaternPC(x, plot = FALSE))
```

We apply the `fitGPmaternPC` function on to each dataset by using `lapply` over
the `thplopit` list of data. The posterior predictive means, standard deviations
and MAP hyperparamters for the GP are returned. As a side effect the posterior
predictive distributions are overlayed with the marker protein profiles for each
subcellular class when the argument `plot = TRUE` is used. Here we do not render
the plots to conform to Biconductor vignette size guidelines.

The prior needs to form a `K*3` matrix (where `K` is the number of subcellular
classes in the data), 

```{r lengthmrk}
(mrkCl <- getMarkerClasses(thplopit[[1]], fcol = "markers"))
```

So for this data we require a `11*3` matrix. Three columns are needed, one for
the prior, one for length-scale amplitude and finally one for the standard
deviation parameters (see `hyppar` in `?fitGP`). For more details see the
manuscript by @bandle.

```{r sethyppar}
K <- length(mrkCl)
pc_prior <- matrix(NA, ncol = 3, K)
pc_prior[seq.int(1:K), ] <- matrix(rep(c(10, 60, 250),
                                       each = K), ncol = 3)
head(pc_prior)
```

Now we have generated these complexity priors we can pass them as an
argument to the `fitGPmaternPC` function. For example,

```{r runhyppar, fig.width=7, fig.height=20}
par(mfrow = c(3,4))
gpParams <- lapply(thplopit,
                   function(x) fitGPmaternPC(x, 
                                             hyppar = pc_prior,
                                             plot = TRUE))
```

The distributions and fit looks sensible for each class so we will proceed with
setting the prior on the weights.

## Setting the prior on the weights

The next step is to set up the matrix Dirichlet prior on the mixing weights.
If `dirPrior = NULL` a default Dirichlet prior is computed see `?bandle`. We
strongly advise you to set your own prior. In "Vignette 1: Getting Started with
BANDLE" we give some suggestions on how to set this and in the below code we try
a few different priors and assess the expectations.

As per Vignette 1, let's try a `dirPrior` as follows,

```{r setweightprior}
set.seed(1)
dirPrior = diag(rep(1, K)) + matrix(0.001, nrow = K, ncol = K)
predDirPrior <- prior_pred_dir(object = thplopit[[1]],
                               dirPrior = dirPrior,
                               q = 15)
```

The mean number of relocalisations is
```{r,}
predDirPrior$meannotAlloc
```

The prior probability that more than `q` differential localisations are
expected is
```{r,}
predDirPrior$tailnotAlloc
```

```{r, fig.height=6, fig.width=6}
hist(predDirPrior$priornotAlloc, col = getStockcol()[1])
```

We see that the prior probability that proteins are allocated to different
components between datasets concentrates around 0. This is what we expect, we
expect subtle changes between conditions for this data. We may perhaps wish to
be a little stricter with the number of differential localisations output by
`bandle` and in this case we could make the off-diagonal elements of the
`dirPrior` smaller. In the below code chunk we test 0.0005 instead of 0.001,
which reduces the number of re-localisations.


```{r try, fig.height=6, fig.width=6}
set.seed(1)
dirPrior = diag(rep(1, K)) + matrix(0.0005, nrow = K, ncol = K)
predDirPrior <- prior_pred_dir(object = thplopit[[1]],
                               dirPrior = dirPrior,
                               q = 15)

predDirPrior$meannotAlloc
predDirPrior$tailnotAlloc
hist(predDirPrior$priornotAlloc, col = getStockcol()[1])
```

Again, we see that the prior probability that proteins are allocated to different
components between datasets concentrates around 0. 

## The bandle function

Now we have computed our `gpParams` and `pcPriors` we can run the main `bandle`
function.

Here for convenience of building the vignette we only run 2 of the triplicates
for each condition and run the `bandle` function for a small number of
iterations to minimise the vignette build-time. Typically we'd recommend you run
the number of iterations (`numIter`) in the $1000$s.

We first subset our data into two objects called `control` and `treatment`
which we subsequently pass to `bandle` along with our priors.

```{r runbandle, message=FALSE}
control <- list(thplopit[[1]], thplopit[[2]])
treatment <- list(thplopit[[3]], thplopit[[4]])

bandleres <- bandle(objectCond1 = control,
                    objectCond2 = treatment,
                    numIter = 50,      # usually 10,000
                    burnin = 5L,        # usually 5,000
                    thin = 1L,          # usually 20
                    gpParams = gpParams,
                    pcPrior = pc_prior,
                    numChains = 1,     # usually >=4
                    dirPrior = dirPrior)
```

A `bandleParams` object is produced
```{r,}
bandleres
```

# Processing and analysing the `bandle` results

Following Vignette 1 we populate the `bandleres` object by calling the
`bandleProcess` function. This may take a few seconds to process.

```{r processbandle}
bandleres <- bandleProcess(bandleres)
```

These slots have now been populated
```{r,}
summary(summaries(bandleres))
```

The `posteriorEstimates` slot gives posterior quantities of interest for
different proteins. The object is of length 2, 
- 1 slot for control 
- 1 slot for treatment

```{r,}
length(summaries(bandleres))
```

We explicitly extract the posterior estimates and protein allocation predictions
as follows

```{r getposteriors}
pe1 <- summaries(bandleres)[[1]]@posteriorEstimates
pe2 <- summaries(bandleres)[[2]]@posteriorEstimates

head(pe1)
```

The full joint probability distribution can be found in the `bandle.joint` slot
e.g. for the control in slot 1 this would be
`summaries(bandleres)[[1]]@bandle.joint` and the treatment in slot 2 this would
be `summaries(bandleres)[[2]]@bandle.joint`.

Let's look at the posterior estimates and allocation predictions found in `pe1`
and `pe2`. Each object is a `data.frame` containing the protein allocations and
associated localisation probabilities for each condition. The 7 columns are

- `bandle.allocation` which contains the the localisation predictions to one of the
subcellular classes that appear in the training data.
- `bandle.probability` is the allocation probability, corresponding to the mean
of the distribution probability.
- `bandle.outlier` is the probability of being an outlier. A high value
indicates that the protein is unlikely to belong to any annotated class (and is
hence considered an outlier).
- `bandle.probability.lowerquantile` and `bandle.probability.upperquantile` are
the upper and lower quantiles of the allocation probability distribution.
- `bandle.mean.shannon` is the Shannon entropy, measuring the uncertainty in the
allocations (a high value representing high uncertainty; the highest value is
the natural logarithm of the number of classes).
- `bandle.differential.localisation` is the differential localisation probability.

We plot the distribution of protein allocations by `bandle` 

```{r barplotalloc, fig.width=8, fig.height=6}
par(mfrow = c(1, 2), oma = c(6,2,2,2))
barplot(table(pe1$bandle.allocation), col = getStockcol()[2],
        las = 2, main = "Control: Protein allocation",
        ylab = "Number of proteins")
barplot(table(pe2$bandle.allocation), col = getStockcol()[2],
        las = 2, main = "Treatment: Protein allocation")
```

The bar plot above tells us for this data `bandle` has allocated the majority of
unlabelled proteins to the nucleus. The allocation result for each condition
(found in `bandle.allocation`) is determined by `bandle` by looking at which
subcellular niche was given the highest probability from the full distribution
e.g. from `bandle.joint`. If we plot the `bandle.probability` (corresponding to
the mean of the distribution) against the protein allocation results we can see
that not all protein allocations are confident, this is why it is important to
threshold when deducing a protein's location.

```{r meanbandleprob, fig.width=8, fig.height=6}
par(mfrow = c(1, 2), oma = c(6,2,2,2))
boxplot(pe1$bandle.probability ~ pe1$ bandle.allocation, 
        col = getStockcol()[2], xlab = "",
        ylab = "BANDLE probability (mean)",
        las = 2, main = "Control: Probability distribution\n by allocation class")
boxplot(pe2$bandle.probability ~ pe1$ bandle.allocation, 
        col = getStockcol()[2], xlab = "", ylab = "",
        las = 2, main = "Treatment: Probability distribution\n by allocation class")
```

## Predicting subcellular location

As mentioned in Vignette 1, it is common to threshold allocation results based
on the posterior probability. Proteins that do not meet the threshold are not
assigned to a subcellular location and left unlabelled (here we use the
terminology "unknown" for consistency with the `pRoloc` package). It is
important not to force proteins to allocate to one of the niches defined here in
the training data, if they have low probability to reside there. We wish to
allow for greater subcellular diversity and to have multiple location, this is
captured essentially in leaving a protein "unlabelled" or "unknown".

We use the `bandlePredict` function to append our results to the original
`MSnSet` datasets. 

```{r predictlocation}
## Add the bandle results to a MSnSet
xx <- bandlePredict(control, 
                    treatment, 
                    params = bandleres, 
                    fcol = "markers")
res_0h <- xx[[1]]
res_12h <- xx[[2]]
```

The BANDLE model combines replicate information within each condition to obtain the
localisation of proteins for each single experimental condition.

The results for each condition are appended to the *first* dataset in the list
of `MSnSets` (for each condition). It is important to familiarise yourself with
the `MSnSet` data structure. To further highlight this in the below code chunk
we look at the `fvarLabels` of each datasets, this shows the column header names
of the `fData` feature data. We see that the first replicate at 0h e.g.
`res_0h[[1]]` has 7 columns with the output of `bandle` e.g.
`bandle.probability`, `bandle.allocation`, `bandle.outlier` etc. (as described
above) appended to the feature data (`fData(res_0h[[1]])`). The second dataset
at 0h i.e. `res_0h[[2]]` does not have this information appended to the feature
data. This is the same for the second condition at 12h post LPS stimulation.

```{r showappended, eval=FALSE}
fvarLabels(res_0h[[1]])
fvarLabels(res_0h[[2]])

fvarLabels(res_12h[[1]])
fvarLabels(res_12h[[2]])
```

To obtain classification results we threshold using a 1% FDR based on the
`bandle.probability` and append the results to the data using the
`getPredictions` function from `MSnbase`.

```{r thresholddata}
## threshold results using 1% FDR
res_0h[[1]] <- getPredictions(res_0h[[1]], 
                              fcol = "bandle.allocation",  
                              scol = "bandle.probability",    
                              mcol = "markers", 
                              t = .99)

res_12h[[1]] <- getPredictions(res_12h[[1]], 
                               fcol = "bandle.allocation",
                               scol = "bandle.probability", 
                               mcol = "markers",      
                               t = .99)
```

A table of predictions is printed as a side effect when running `getPredictions` function. 

In addition to thresholding on the `bandle.probability` we can threshold
based on the `bandle.outlier` i.e. the probability of being an outlier. A high value
indicates that the protein is unlikely to belong to any annotated class (and is
hence considered an outlier). We wish to assign proteins to a subcellular niche 
if they have a high `bandle.probability` and also a low `bandle.outlier` probability.
This is a nice way to ensure we keep the most high confidence localisations. 

In the below code chunk we use first create a new column called `bandle.outlier.t`
in the feature data which is `1 - outlier probability`. This allows us then to use
`getPredictions` once again and keep only proteins which meet both the 0.99
threshold on the `bandle.probability` and the `bandle.outlier`.

Note, that running `getPredictions` appends the results to a new feature data
column called `fcol.pred`, please see `?getPredictions` for the documentation.
As we have run this function twice, our column of classification results
are found in `bandle.allocation.pred.pred`.

```{r threshold2}
## add outlier probability
fData(res_0h[[1]])$bandle.outlier.t <- 1 -  fData(res_0h[[1]])$bandle.outlier
fData(res_12h[[1]])$bandle.outlier.t <- 1 -  fData(res_12h[[1]])$bandle.outlier

## threshold again, now on the outlier probability
res_0h[[1]] <- getPredictions(res_0h[[1]], 
                              fcol = "bandle.allocation.pred",  
                              scol = "bandle.outlier.t",    
                              mcol = "markers", 
                              t = .99)

res_12h[[1]] <- getPredictions(res_12h[[1]], 
                               fcol = "bandle.allocation.pred",
                               scol = "bandle.outlier.t", 
                               mcol = "markers",      
                               t = .99)
```

Let's append the results to the second replicate (by default they are appended
to the first only, as already mentioned above). This allows us to plot each
dataset and the results using `plot2D`.

```{r appendtosecond}
## Add results to second replicate at 0h
res_alloc_0hr <- fData(res_0h[[1]])$bandle.allocation.pred.pred
fData(res_0h[[2]])$bandle.allocation.pred.pred <- res_alloc_0hr

## Add results to second replicate at 12h
res_alloc_12hr <- fData(res_12h[[1]])$bandle.allocation.pred.pred
fData(res_12h[[2]])$bandle.allocation.pred.pred <- res_alloc_12hr
```

We can plot these results on a PCA plot and compare to the original subcellular
markers.

```{r plotmyres, fig.height=14, fig.width=5}
par(mfrow = c(5, 2))

plot2D(res_0h[[1]], main = "Unstimulated - replicate 1 \n subcellular markers", 
       fcol = "markers")
plot2D(res_0h[[1]], main = "Unstimulated - replicate 1 \nprotein allocations (1% FDR)", 
       fcol = "bandle.allocation.pred.pred")

plot2D(res_0h[[2]], main = "Unstimulated - replicate 2 \nsubcellular markers", 
       fcol = "markers")
plot2D(res_0h[[2]], main = "Unstimulated - replicate 2 \nprotein allocations (1% FDR)", 
       fcol = "bandle.allocation.pred.pred")

plot2D(res_0h[[1]], main = "12h LPS - replicate 1 \nsubcellular markers", 
       fcol = "markers")
plot2D(res_0h[[1]], main = "12h LPS - replicate 1 \nprotein allocations (1% FDR)", 
       fcol = "bandle.allocation.pred.pred")

plot2D(res_0h[[2]], main = "12h LPS - replicate 2 \nsubcellular markers", 
       fcol = "markers")
plot2D(res_0h[[2]], main = "12h LPS - replicate 2 \nprotein allocations (1% FDR)", 
       fcol = "bandle.allocation.pred.pred")

plot(NULL, xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
addLegend(res_0h[[1]], where = "topleft", cex = .8)
```


## Differential localisation

The differential localisation probability tells us which proteins are most
likely to *differentially localise*, that exhibit a change in their steady-state
subcellular location. Quantifying changes in protein subcellular location
between experimental conditions is challenging and Crook et al [@bandle] have
used a Bayesian approach to compute the probability that a protein
differentially localises upon cellular perturbation, as well quantifying the
uncertainty in these estimates. The differential localisation probability is
found in the `bandle.differential.localisation` column of the `bandleParams`
output.

```{r numtransloc}
diffloc_probs <- pe1$bandle.differential.localisation
```

If we take a 5% FDR and examine how many proteins get a differential probability
greater than 0.95 we find there are 
`r length(which(diffloc_probs[order(diffloc_probs, decreasing = TRUE)] > 0.99))` 
proteins above this threshold.

```{r cutoffres}
length(which(diffloc_probs[order(diffloc_probs, decreasing = TRUE)] > 0.95))
```

On a rank plot we can see the distribution of differential probabilities.
                                                                                                                                                                                                                                                              
```{r extractdifflocp, fig.height=6, fig.width=7}
plot(diffloc_probs[order(diffloc_probs, decreasing = TRUE)],
     col = getStockcol()[2], pch = 19, ylab = "Probability",
     xlab = "Rank", main = "Differential localisation rank plot")
```

This indicated that most proteins are not differentially localised and there are a
few hundred confident differentially localised proteins of interest. 

## Estimating uncertainty

One advantage of using Bayesian methods over classic machine learning is the
ability to quantify the uncertainty in our estimates. There are many ways to do
this, as discussed in "Vignette 1: Getting Started with BANDLE". In the below
code chunk we use the `binomialDiffLocProb` function to obtain credible
intervals from the binomial distribution and then extract a probability
estimate for the differential localisation. 

Please note, that in interest of time and for the purpose of demonstration we 
set `nsample = 500` and thus only return 500 samples of the binomial distribution.
In practice the minimum recommended number of samples is 5000.


```{r diffloc_binom}
set.seed(1)
bin_t <- binomialDiffLocProb(params = bandleres, top = 500,
                             nsample = 500, decreasing = TRUE)
```

As we have a large number of proteins as candidates we have chosen to threshold
on the interval to reduce the number of differential localisations.

```{r get_pe}
qt <- apply(bin_t, 1, function(x) quantile(x, .025))
```

This leaves us with `r sum(qt > 0.95)` proteins to investigate.

```{r candidates}
candidates <- names(which(qt > 0.95))
head(candidates)
```

### Add the results to the MSnSet

Let's add the results to each replicate in the `MSnSet`s.
The reason for doing this is so that later on when we wish
to visulalise the data we have the information readily
accessible to make use of the functions in the `pRoloc`
package.

Let's double check all datasets have the same proteins,
```{r chkallsame}
all(featureNames(res_0h[[1]]) == featureNames(res_0h[[2]]))
all(featureNames(res_0h[[1]]) == featureNames(res_12h[[1]]))
all(featureNames(res_12h[[1]]) == featureNames(res_12h[[2]]))
```

Now let's add the differential location estimates,

```{r addtomsn}
dl.estimate <- qt[candidates]
fn <- featureNames(control[[1]])
cmn <- fn %in% names(dl.estimate)


## Add results to the 0h time-point (control)
for (i in seq(res_0h)) {
    ## create column called "dl.estimate" in the data
    mcol <- "dl.estimate"
    fData(res_0h[[i]])[, mcol] <- NA
    fData(res_0h[[i]])[cmn, mcol] <- dl.estimate
    ## create column called "dl.candidate" in the data
    mcol <- "dl.candidate"
    fData(res_0h[[i]])[, mcol] <- "unknown"
    fData(res_0h[[i]])[cmn, mcol] <- "DL candidate"
}


## Add results to the 12h time-point (treatment)
for (i in seq(res_12h)) {
    ## create column called "dl.estimate" in the data
    mcol <- "dl.estimate"
    fData(res_12h[[i]])[, mcol] <- NA
    fData(res_12h[[i]])[cmn, mcol] <- dl.estimate
    ## create column called "dl.candidate" in the data
    mcol <- "dl.candidate"
    fData(res_12h[[i]])[, mcol] <- "unknown"
    fData(res_12h[[i]])[cmn, mcol] <- "DL candidate"
}
```

In the next section we can visualise these results.


# Visualising differential localisation


There are several different ways we can visualise the output of `bandle`. Now we
have our set of candidates we can subset `MSnSet` datasets and plot the the
results.

To subset the data, 
```{r alluvial, warning=FALSE, message=FALSE}
msnset_cands <- list(res_0h[[1]][candidates, ], 
                     res_12h[[1]][candidates, ])
```

We can visualise this as a `data.frame` too for ease,

```{r dataframeres}
# construct data.frame
df_cands <- data.frame(
    fData(msnset_cands[[1]])[, c("dl.estimate", "bandle.allocation.pred.pred")],
    fData(msnset_cands[[2]])[, "bandle.allocation.pred.pred"])
colnames(df_cands) <- c("dl.estimate", "0hr_location", "12h_location")

# order by highest differential localisation estimate
df_cands <- df_cands %>% arrange(desc(dl.estimate))
head(df_cands)
```

## Alluvial plots

We can now plot this on an alluvial plot to view the changes in subcellular
location. The class label is taken from the column called
`"bandle.allocation.pred.pred"` which was deduced above by thresholding on the
posterior and outlier probabilities before assigning BANDLE's allocation
prediction.

```{r plotres, fig.height=10, fig.width=8}
## set colours for organelles and unknown
cols <- c(getStockcol()[seq(mrkCl)], "grey")
names(cols) <- c(mrkCl, "unknown")

## plot
alluvial <- plotTranslocations(msnset_cands, 
                               fcol = "bandle.allocation.pred.pred", 
                               col = cols)
alluvial + ggtitle("Differential localisations following 12h-LPS stimulation")
```

To view a table of the translocations, we can call the argument `table = TRUE`,

```{r tbllocs}
tbl <- plotTranslocations(msnset_cands, 
                               fcol = "bandle.allocation.pred.pred", 
                               col = cols, 
                               table = TRUE)
tbl
```

Although this example analysis is limited compared to that of @thplopit, we do
see similar trends inline with the results seen in the paper. For examples, we
see a large number of proteins translocating between organelles that are
involved in the secretory pathway. We can further examine these cases by
subsetting the datasets once again and only plotting proteins that involve
localisation with these organelles. Several organelles are known to be involved 
in this pathway, the main ones, the ER, Golgi (and plasma membrane). 

Let's subset for these proteins,
```{r plotlysos, fig.height=10, fig.width=8}
secretory_prots <- unlist(lapply(msnset_cands, function(z) 
    c(which(fData(z)$bandle.allocation.pred.pred == "Golgi Apparatus"),
      which(fData(z)$bandle.allocation.pred.pred == "Endoplasmic Reticulum"),
      which(fData(z)$bandle.allocation.pred.pred == "Plasma Membrane"),
      which(fData(z)$bandle.allocation.pred.pred == "Lysosome"))))
secretory_prots <- unique(secretory_prots)

msnset_secret <- list(msnset_cands[[1]][secretory_prots, ], 
                    msnset_cands[[2]][secretory_prots, ])

secretory_alluvial <- plotTranslocations(msnset_secret, 
                                         fcol = "bandle.allocation.pred.pred", 
                                         col = cols)
secretory_alluvial + ggtitle("Movements of secretory proteins")
```

## Protein profiles

In the next section we see how to plot proteins of interest.Our differential
localisation candidates can be found in `df_cands`,

```{r plotdfprof}
head(df_cands)
```

Let's take the first protein as an example; protein with accession B2RUZ4. It
has a high differential localisation score and it's steady state localisation in
the control is predicted to be lysosomal and in the treatment condition at 12
hours-LPS it is predicted to localise to the plasma membrane. This fits with the
information we see on Uniprot which tells us it is Small integral membrane
protein 1 (SMIM1).

In the below code chunk we plot the protein profiles of all proteins that were
identified as lysosomal from BANDLE in the control and then overlay SMIM1. We do
the same at 12hrs post LPS with all plasma membrane proteins.

```{r protprof, fig.height=7, fig.width=8}
par(mfrow = c(2, 1))

## plot lysosomal profiles
lyso <- which(fData(res_0h[[1]])$bandle.allocation.pred.pred == "Lysosome")
plotDist(res_0h[[1]][lyso], pcol = cols["Lysosome"], alpha = 0.04)
matlines(exprs(res_0h[[1]])["B2RUZ4", ], col = cols["Lysosome"], lwd = 3)
title("Protein SMIM1 (B2RUZ4) at 0hr (control)")

## plot plasma membrane profiles
pm <- which(fData(res_12h[[1]])$bandle.allocation.pred.pred == "Plasma Membrane")
plotDist(res_12h[[1]][pm], pcol = cols["Plasma Membrane"], alpha = 0.04)
matlines(exprs(res_12h[[1]])["B2RUZ4", ], col = cols["Plasma Membrane"], lwd = 3)
title("Protein SMIM1 (B2RUZ4) at 12hr-LPS (treatment)")
```

We can also visualise there on a PCA or t-SNE plot.

```{r plotpcacands, fig.height=6, fig.width=9}
par(mfrow = c(1, 2))
plot2D(res_0h[[1]], fcol = "bandle.allocation.pred.pred",
       main = "Unstimulated - replicate 1 \n predicted location")
highlightOnPlot(res_0h[[1]], foi = "B2RUZ4")

plot2D(res_12h[[1]], fcol = "bandle.allocation.pred.pred",
       main = "12h-LPS - replicate 1 \n predicted location")
highlightOnPlot(res_12h[[1]], foi = "B2RUZ4")
```


# Session information

All software and respective versions used to produce this document are listed below.
```{r sessionInfo}
sessionInfo()
```

# References {-}